PTFO 8-8695: Two Stars, Two Signals, No Planet

### 2020-06-12 paper_review/response_20200605.txt

******************************************
REFEREE COMMENT
----------

Introduction: I do not think this is youngest hot Jupiter. That should be true
just for transiting systems, and it was the youngest at discovery, but the two
RV planets from ESPaDOnS are younger (Donati et al. 2016 and Yu et al. 2017).
Both are hot Jupiters and in Taurus. I believe the age of Tap 26 is more
ambiguous due to age spreads (multiple groups?) in Taurus, but the age of V830
Tau is certainly <3 Myr.

RESPONSE
----------

> Thank you for this correction. We have qualified the relevant sentence as the
> youngest known *transiting* hot Jupiter, and added a footnote to the V830
> Tau or TAP 26 results.

@@

RESPONSE
----------
> We've removed it -- thank you for the suggestion.

******************************************
REFEREE COMMENT
----------

Section 3/4: This deals mostly with identification of the second star. However,
the second star appears to be an assumption of section 3.2. If the periods come
from two stars that could be separated (e.g., wide binaries), it would have
been best to do so rather than try to model both by including fs and fl terms.
So it is odd that the binary search is after the light curve model in the
paper. It might make more narrative sense to move the light curve model to
after section 4, since 4 motivates 3.2.

RESPONSE
----------
> While we see the referee's point, we prefer the narrative structure of
> presenting the TESS data first, and getting into considerations of stellar
> binarity, both resolved and unresolved, afterwards.
>
> The reason is that the multiple signals come from TESS data processed at the
> ``default'' level. From a story-telling perspective, why would we worry about
> binarity if we didn't already know there were multiple photometric signals?

@@

******************************************
REFEREE COMMENT
----------

Section 5.2: That clear offset at (Pl-Ps)/Ps doesn't seem like a coincidence,
and the paper reads like the authors agree, but couldn't explain it. I'm not
sure I can help much here, although it is interesting. Could this be that both
stars are dippers and one turns off while the other turns on by coincidence?
Spots and hot-spot clusters can shift in phase as the magnetic fields realign,
which creates a new spot cluster at the same period but a different phase. If
it is circumstellar material, a tight binary could be shepherding the material,
but I think the systems would have to be very tight for that, and it could not
last for such a long time.

I think this can be assessed statistically; given the scatter in those points,
the odds of lining up to any given line is small. But without a physical
explanation this is hard to test. Probably this requires looking at many
dippers as an ensemble. It is not required for the primary analysis of this
paper, so I will defer to the authors if they want to dig deeper on this.

RESPONSE
----------
> Indeed, we are not sure whether or not it is a coincidence, but we
> thought it was interesting enough to mention, even though we did not
> have an interpretation.
>
> It seems unlikely in our view that both stars are showing dips.
> Our interpretation of the two smooth photometric signals is that one
> star is rotating at 11.98 hr, and the other star is rotating at
> 10.76 hr.
>
> All of the dips occur with a periodicity of 10.76 hr. (If we make
> Figure 6 but instead phasing by P_long, it yields a mess of points).
> Therefore all the dips are presumably associated with the star that
> has that rotation period.

@@

******************************************
REFEREE COMMENT
----------

Section 5.3.2: The difference between the fraction of transient flux dip stars
and the binary fraction is more significant than it appears to the authors.
High mass ratio binaries would not be included, as they are too faint to detect
the second period. Further, at the distance of Upper Sco (~150pc) and assuming
Stauffer et al. 2017 would notice/separate two stars wider than a K2 pixel
(4"), the presence of two periods would not include binaries wider than 600pc.
Eyeballing the Winters et al. 2019 distribution suggests ~20% of binaries are
beyond this. Stauffer et al. 2017 also probably checked against existing
imaging data for obvious binaries down to 1-2", so likely even more are getting
cut.

RESPONSE
----------
> We have updated the text to include this point -- thank you for
> making it.
> 
> We have also updated the text to reflect our hesitation about
> whether or not the phenomenon is associated with binarity, primarily
> because multiple photometric periods do not necessarily imply
> binarity.

@@

******************************************
REFEREE COMMENT
----------

Section 5.4: The authors conclude this subsection with a reasonable argument
that this isn't a protoplanet. Does this same argument not rule out gas/debris
at the Keplerian coronation radius? What could maintain such clouds for
years/decades and then have them jump phase but at the same period? It's hard
to do this if it is not somehow physically attached to the star in some what
(e.g., spots can do this).

RESPONSE
----------
> We have modified the last paragraphs of Section 5.4 to reflect an
> updated preference that we came to after submission, for the clumpy
> torus model over the spot occultation model. The reasons
> for the preference are given in the text: 1) the spot occultation
> model seems more geometrically contrived than we initially thought,
> and 2) the Halpha excess seen by Johns-Krull+16 (and also in
> unpublished observations by Stauffer of scallop shells) seems likely
> to require hydrogen in orbit near the corotation radius.

@@

A few additional changes are noted in the "List of Changes" appended at the end
of the manuscript.

We modified the text throughout to clarify that Stauffer+ 2017 favored the
explanation of a clumpy torus of gas and dust at the Keplerian co-rotation
radius, rather than our initial (incomplete) description of "clouds of gas" at
the same location.

We also modified the labels of the axes in Figure 5a for clarity.

We have also added a few citations that we omitted in the original
manuscript.

### 2020-06-12 paper/ms.tex
wrap up the referee response

We wish \ptfob\ were a planet. It would be quite exceptional. It would
be the youngest known \added{transiting} hot Jupiter \citep{van_eyken_ptf_2012}, orbiting
a T Tauri star in the Orion-OB1a cluster.  It would have the shortest
orbital period of any hot Jupiter.  With such a short period, it would
probably be filling its Roche lobe, and actively losing mass to its
host star.  Not only that, but the rapidly-rotating host star is
probably oblate enough to \replaced{torque the planet's orbit into and out of
the transiting configuration}{cause the planet's orbit to precess into and out of the 
transiting configuration} on a timescale of years
\citep{barnes_measurement_2013,ciardi_followup_2015,kamiaka_revisiting_2015}.

@@

We acquired the RUWE for each of the
149 members in 25$\,$Ori-1 identified by
identified by \citet{kounkel_apogee2_2018}.
We then queried the
CDIPS light curve database at MAST \citep{bouma_cluster_2019} to find
the subset of members that were at least as variable as \ptfo.  We
measured the variability amplitude by taking the difference between
the $95^{\rm th}$ and $5^{\rm th}$ percentiles of the flux
measurements.  This yielded 30 stars of equal or greater variability.
The lower panel of Figure~\ref{fig:gaia} shows the RUWE of
these stars as
a function of stellar brightness.  \ptfo\ is at the 93$^{\rm
	rd}$ percentile of equally-variable stars within the
25$\,$Ori-1 group.  
Two of the thirty stars
with variability amplitudes greater than 9.7\% showed higher RUWE.
One was CVSO~35, which has a TESS light curve that varies by 2
magnitudes.  CVSO~35 also shows a strong Wide-field Infrared Survey
Exoplorer (WISE, \citealt{wright_WISE_2010}) IR excess as well as a
10$\mu$m silicate emission feature \citep{mauco_herschel_2018}.  The
other star was GAIA DR2 3222210363837122048.

@@

Second, approximately 40-50\% of the transient flux dip stars
discovered in $\rho$~Oph and Upper~Sco show two Lomb-Scargle periods,
and so are apparently binaries
\citep[][Table~1]{stauffer_orbiting_2017}.  This is higher than the
main-sequence companion fraction of ${\rm
CF}_{0.1\text{-}0.5\,M_\odot}^{\rm MS} = 33 \pm 5\%$
\citep{henry_solar_2006,duchene_stellar_2013,winters_solar_2019}.
Low-mass pre-main-sequence stars however have been shown to companion
fractions up to twice as high in dispersed clusters such as Upper~Sco
and Taurus \citep{kraus_mapping_2008,kraus_mapping_2011}.  A
high-resolution imaging survey would be interesting, to determine
whether the transient flux dip stars truly have distinct
population-level binarity properties relative to other young low-mass
stars.

@@

The explanations that are not yet ruled out include {\it (i)}
transiting clumps of \added{dust or }gas at the Keplerian co-rotation radius; {\it
(ii)} transits of enshrouded protoplanets; {\it (iii)} occultations of
starspots by an optically thick disk.  The first and last explanations
have added appeal because they are flexible enough to explain not only
the transient and persistent-dip M-dwarfs, but also the ``scallop
shell'' M-dwarfs \citep{stauffer_orbiting_2017}.  \replaced{Despite this appeal,
the possibility of distinct mechanisms explaining these distinct
variability classes remains open.}{
The model of starspot occultations by an optically thick disk \citep{zhan_complex_2019}
is a qualitative inverse of the torus of circumstellar material:
the difference is whether to place the ``blobs'' that
provide the needed brightness contrast on the
surface of the star, or at the co-rotation radius.
}

\replaced{The evolution of \ptfo\ over}{Observations of \ptfo\ from} the past decade
could offer important hints.
\deleted{Specifically, }\ptfo's transition between having none, one, and
multiple dips per cycle seems important \added{(Figure~\ref{fig:o_minus_c})}.  It strains the ``enshrouded
protoplanet'' interpretation, because there are no known processes
that cause a planet's orbital phase to jump.  \deleted{The dips would then need
to be caused by material that was somehow disrupted from the planet,
but somehow remained co-orbital for an extended duration. This seems
implausible.}
\added{
The H$\alpha$
excess observed by \citet{johnskrull_h_2016} also seems like an
important piece of evidence.
The model with circumstellar optically thick hydrogen
gas near the co-rotation radius easily explains this excess,
while the starspot occultation model does not.
}

>>>>

We wish \ptfob\ were a planet. It would be quite exceptional. It would
be the youngest known \added{transiting\footnote{Two younger hot
Jupiters have been reported around V830 Tau and TAP 26
\citep{donati_hot_2016,donati_hot_2017,yu_hot_2017}.}} hot Jupiter \citep{van_eyken_ptf_2012}, orbiting
a T Tauri star in the Orion-OB1a cluster.  It would have the shortest
orbital period of any hot Jupiter.  With such a short period, it would
probably be filling its Roche lobe, and actively losing mass to its
host star.  Not only that, but the rapidly-rotating host star is
probably oblate enough to \replaced{torque the planet's orbit into and out of
the transiting configuration}{cause the planet's orbit to precess into and out of the 
transiting configuration} on a timescale of years
\citep{barnes_measurement_2013,ciardi_followup_2015,kamiaka_revisiting_2015}.

@@

We acquired the RUWE for each of the
149 members in 25$\,$Ori-1 identified by
identified by \citet{kounkel_apogee2_2018}.
We then queried the
CDIPS light curve database at MAST \citep{bouma_cluster_2019} to find
the subset of members that were at least as variable as \ptfo.  We
measured the variability amplitude by taking the difference between
the $95^{\rm th}$ and $5^{\rm th}$ percentiles of the flux
measurements.  This yielded 30 stars of equal or greater variability.
The lower panel of Figure~\ref{fig:gaia} shows the RUWE of
these stars as
a function of stellar brightness.  \ptfo\ is at the 93$^{\rm
	rd}$ percentile of equally-variable stars within the
25$\,$Ori-1 group.  
Two of the thirty stars
with variability amplitudes greater than 9.7\% showed higher RUWE.
One was CVSO~35, which has a TESS light curve that varies by 2
magnitudes, and shows both a strong Wide-field Infrared Survey
Exoplorer (WISE, \citealt{wright_WISE_2010}) IR excess and also a
10$\mu$m silicate emission feature \citep{mauco_herschel_2018}.  The
other star was GAIA DR2 3222210363837122048.

@@

Second, approximately 40-50\% of the transient flux dip stars
discovered in $\rho$~Oph and Upper~Sco show two Lomb-Scargle periods,
and so \replaced{are apparently}{could be} binaries
\citep[][Table~1]{stauffer_orbiting_2017}.  This is higher than the
main-sequence companion fraction of ${\rm
CF}_{0.1\text{-}0.5\,M_\odot}^{\rm MS} = 33 \pm 5\%$\added{, and so it
could be an indication that the transient flux dip phenomenon is
linked to stellar binarity}
\citep{henry_solar_2006,duchene_stellar_2013,winters_solar_2019}.
\added{This fraction might be even more significant given the
	selection bias against high
mass ratio binaries, in which
the second star would be too faint to detect a second period.}

\added{However, the association between transient flux dips and binarity is
not iron-clad.  Firstly, multiple photometric periods do not necessarily
imply binarity. For instance, circumstellar material could exist
around single stars inside, at, or wide of co-rotation.  Additionally, } low-mass pre-main-sequence stars
\deleted{however} have been shown to \added{have} companion fractions
up to twice as high in dispersed clusters such as Upper~Sco and Taurus
\citep{kraus_mapping_2008,kraus_mapping_2011}.  A high-resolution
imaging survey would be interesting, to determine whether the
transient flux dip stars truly have distinct population-level binarity
properties relative to other young low-mass stars.

@@

The explanations that \replaced{are not yet}{have not yet been} ruled out include {\it (i)}
transiting clumps of \added{dust or }gas at the Keplerian co-rotation radius; {\it
(ii)} transits of enshrouded protoplanets; {\it (iii)} occultations of
\replaced{starspots}{hot and cold spots on the stellar surface} by an optically thick disk.  The first and last explanations
have added appeal because they are flexible enough to explain not only
the transient and persistent-dip M-dwarfs, but also the ``scallop
shell'' M-dwarfs \citep{stauffer_orbiting_2017}.  \replaced{Despite this appeal,
the possibility of distinct mechanisms explaining these distinct
variability classes remains open.}{
The model of hot and coldspot occultations by an optically thick disk
\citep{zhan_complex_2019} is a qualitative inverse of the torus of
circumstellar material: the difference is whether to place the
``blobs'' that provide the needed brightness contrast on the surface
of the star, or at the co-rotation radius.  In the spot occultation
model, the observing geometry would require an inner disk to be
projected into a narrow band across the stellar surface, in order to
produce variations with sufficiently short timescales.  The outer
disk would need either to not be present, or to be misaligned enough,
and narrow enough, that it did not fall along the line of sight. These
combined requirements seem geometrically contrived.
}

\replaced{The evolution of \ptfo\ over}{Observations of \ptfo\ from} the past decade
could offer additional hints.
\deleted{Specifically, }\ptfo's transition between having none, one, and
multiple dips per cycle seems important \added{(Figure~\ref{fig:o_minus_c})}.  It strains the ``enshrouded
protoplanet'' interpretation, because there are no known processes
that cause a planet's orbital phase to jump.  \deleted{The dips would then need
to be caused by material that was somehow disrupted from the planet,
but somehow remained co-orbital for an extended duration. This seems
implausible.}
\added{
The H$\alpha$
excess observed by \citet{johnskrull_h_2016} also seems like an
important piece of evidence.
The model with circumstellar optically thick hydrogen
gas near the co-rotation radius readily explains this excess.
}
